---
title: "JJ - Forecast evaluation"
output: html_document
date: "2026-09-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import settings
```{r config}
library(RcppTOML)

config <- parseTOML("config.toml")

COUNTRIES <- config$COUNTRIES
LOOKBACK <- config$LOOKBACK
N_SAMPLES <- config$N_SAMPLES
FORECAST_HORIZONS <- config$FORECAST_HORIZONS
QUANTILES <- config$QUANTILES
FORECAST_START_DATE <- as.Date(config$FORECAST_START_DATE)
INDICATORS <- config$INDICATORS
RESULTS_PATH <- config$RESULTS_PATH
BISTRO_ROOT <- config$BISTRO_ROOT
DATA_PATHS <- config$DATA_PATHS
```

# Import data
```{r results}
files <- list.files(RESULTS_PATH, pattern = "\\.csv$", full.names = TRUE)
results <- lapply(files, read.csv)
names(results) <- tools::file_path_sans_ext(basename(files))

cpi <- read.csv(DATA_PATHS$cpi)
ipi <- read.csv(DATA_PATHS$ipi)
```

# Evaluation setup

Select multiple indicators here when required. Each indicator is evaluated
independently: scores, tables, and plots are produced separately and are never
averaged or pooled across indicators.

```{r evaluation-setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(scoringRules)

SELECTED_COUNTRIES <- COUNTRIES
SELECTED_HORIZONS <- c(1, 12)
SELECTED_INDICATORS <- c("cpi", "ipi")
SELECTED_MODELS <- c("chronos2", "gas", "qar", "sundial", "timesfm")
# SELECTED_MODELS <- c("historical")
PLOT_COUNTRY <- SELECTED_COUNTRIES[1]

INDICATOR_DATA <- list(cpi = cpi, ipi = ipi)
PLOT_DIR <- "plots"
source(file.path(PLOT_DIR, "theme.R"))
source(file.path(PLOT_DIR, "model_comparison.R"))
source(file.path(PLOT_DIR, "wis_decomposition.R"))
source(file.path(PLOT_DIR, "calibration.R"))
source(file.path(PLOT_DIR, "murphy.R"))

stopifnot(all(SELECTED_INDICATORS %in% names(INDICATOR_DATA)))
```

# Evaluation by indicator

Every iteration below is a separate evaluation problem. The named
`evaluation_results` list preserves the indicator boundary for downstream use;
there is no combined score across indicators.

```{r evaluate-indicators, results='hide'}
TAU <- sort(as.numeric(QUANTILES))
LOWER_TAUS <- TAU[TAU < 0.5]
K <- length(LOWER_TAUS)

model_name_for <- function(result_name, indicator_suffix) {
  substr(result_name, 1, nchar(result_name) - nchar(indicator_suffix))
}

load_indicator_forecasts <- function(indicator, horizon) {
  indicator_suffix <- paste0("_", indicator, "_forecasts")
  candidates <- names(results)[endsWith(names(results), indicator_suffix)]
  selected <- candidates[
    vapply(candidates, model_name_for, character(1), indicator_suffix) %in%
      SELECTED_MODELS
  ]

  bind_rows(lapply(selected, function(result_name) {
    df <- results[[result_name]]
    df$model <- model_name_for(result_name, indicator_suffix)
    df
  })) %>%
    filter(
      country %in% SELECTED_COUNTRIES,
      horizon == !!horizon,
      !is.na(actual)
    ) %>%
    mutate(
      origin = as.Date(origin),
      target_date = origin %m+% months(!!horizon)
    )
}

evaluate_indicator <- function(indicator, horizon) {
  forecasts <- load_indicator_forecasts(indicator, horizon)
  historical <- INDICATOR_DATA[[indicator]] %>%
    select(TIME_PERIOD, all_of(PLOT_COUNTRY)) %>%
    rename(date = TIME_PERIOD, value = all_of(PLOT_COUNTRY)) %>%
    mutate(date = as.Date(date))

  y <- forecasts$actual
  median_ae <- abs(y - forecasts[["q0.5"]])
  is_mat <- sapply(LOWER_TAUS, function(tau) {
    alpha <- 2 * tau
    lower <- forecasts[[paste0("q", tau)]]
    upper <- forecasts[[paste0("q", 1 - tau)]]
    (alpha / 2) * scoringRules::ints_quantiles(
      y, lower, upper, target_coverage = 1 - alpha
    )
  })
  disp_mat <- sapply(LOWER_TAUS, function(tau) {
    alpha <- 2 * tau
    lower <- forecasts[[paste0("q", tau)]]
    upper <- forecasts[[paste0("q", 1 - tau)]]
    (alpha / 2) * (upper - lower)
  })

  forecasts <- forecasts %>%
    mutate(
      wis = (0.5 * median_ae + rowSums(is_mat)) / (K + 0.5),
      sharpness = rowSums(disp_mat) / (K + 0.5),
      calibration = wis - sharpness
    ) %>%
    rowwise() %>%
    mutate(
      pit = approx(
        x = c_across(all_of(paste0("q", TAU))),
        y = TAU, xout = actual, rule = 2
      )$y
    ) %>%
    ungroup()

  scores_summary <- forecasts %>%
    group_by(model) %>%
    summarise(
      wis = mean(wis), sharpness = mean(sharpness),
      calibration = mean(calibration), n = n(), .groups = "drop"
    ) %>%
    arrange(wis)

  coverage <- bind_rows(lapply(LOWER_TAUS, function(tau) {
    alpha <- 2 * tau
    lower <- forecasts[[paste0("q", tau)]]
    upper <- forecasts[[paste0("q", 1 - tau)]]
    forecasts %>%
      mutate(inside = actual >= lower & actual <= upper) %>%
      group_by(model) %>%
      summarise(empirical_coverage = mean(inside), .groups = "drop") %>%
      mutate(nominal_coverage = 1 - alpha)
  }))

  list(
    forecasts = forecasts,
    historical = historical,
    scores_summary = scores_summary,
    coverage = coverage,
    murphy = murphy_scores(forecasts)
  )
}

evaluation_keys <- expand.grid(
  indicator = SELECTED_INDICATORS, horizon = SELECTED_HORIZONS,
  stringsAsFactors = FALSE
)

evaluation_results <- setNames(
  Map(evaluate_indicator, evaluation_keys$indicator, evaluation_keys$horizon),
  paste(evaluation_keys$indicator, evaluation_keys$horizon, sep = "_h")
)
```

Each indicator-horizon pair below is evaluated and reported fully separately:
no scores, tables, or plots ever mix horizons.

```{r report-indicators, results='asis'}
for (i in seq_len(nrow(evaluation_keys))) {
  indicator <- evaluation_keys$indicator[i]
  horizon <- evaluation_keys$horizon[i]
  result <- evaluation_results[[paste0(indicator, "_h", horizon)]]
  forecasts <- result$forecasts
  historical <- result$historical
  scores_summary <- result$scores_summary
  coverage <- result$coverage

  cat("\n## ", toupper(indicator), " - horizon ", horizon, "\n\n", sep = "")
  cat("All metrics below use only ", toupper(indicator),
      " forecasts at horizon ", horizon,
      " and are not combined with other indicators or horizons.\n\n", sep = "")

  print(plot_model_comparison(
    forecasts, historical, indicator, horizon, PLOT_COUNTRY
  ))
  print(scores_summary)
  print(plot_wis_decomposition(
    scores_summary, indicator, horizon, SELECTED_COUNTRIES
  ))
  print(plot_pit(
    forecasts, indicator, horizon, SELECTED_COUNTRIES
  ))
  print(plot_coverage(
    coverage, indicator, horizon, SELECTED_COUNTRIES
  ))
  print(plot_murphy(
    result$murphy, indicator, horizon, SELECTED_COUNTRIES
  ))
}
```
